---
title: "Moderation Analysis: Schooling × Fossil-fuel share of Electricity"
author: "Yunting Du"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    theme: readable
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  fig.width = 10,
  fig.height = 7
)
library(here)
dir.create(here("output"), showWarnings = FALSE, recursive = TRUE)
```
# Packages
```{r}
library(dplyr)
library(readr)
library(tidyr)
library(ggplot2)
library(car)
library(sandwich)
library(lmtest)
library(stargazer)
library(DescTools)
library(interactions)
library(knitr)
library(here)
```

# Paths
```{r}
path_energy <- here("owid-energy-data.csv")
path_co2    <- here("owid-co2-data.csv")
path_school <- here("average_years_of_schooling.csv")
path_gdp    <- here("GDP_per_capita.csv")

target_year <- 2023
```

```{r}
energy <- read_csv(path_energy)
co2    <- read_csv(path_co2)
school <- read_csv(path_school)
gdp    <- read_csv(path_gdp)
```
# Data cleaning
## Fossilfuel share data
```{r}
fossil_share_elec <- read_csv(path_energy, show_col_types = FALSE)

fossil_share_elec_2023 <- fossil_share_elec %>%
filter(nchar(iso_code) == 3, year == target_year) %>%
select(country, iso_code, year, fossil_share_elec)

head(fossil_share_elec_2023)

write_csv(
fossil_share_elec_2023,
here("output", paste0("owid-fossil-data-cleaned-", target_year, ".csv"))
)
```
## CO2 per capita data
```{r}
co2pc <- read_csv(path_co2, show_col_types = FALSE)

co2_per_capita_2023 <- co2pc %>%
filter(nchar(iso_code) == 3, year == target_year) %>%
select(country, iso_code, year, co2_per_capita)

head(co2_per_capita_2023)

write_csv(
co2_per_capita_2023,
here("output", paste0("owid-co2pc-data-cleaned-", target_year, ".csv"))
)
```
## Mean years of schooling data
```{r}
schooling_2023 <- read_csv(path_school, show_col_types = FALSE) %>%
transmute(
country   = Entity,
iso_code  = iso_code,
year      = year,
schooling = as.numeric(schooling)
) %>%
filter(nchar(iso_code) == 3, year == target_year)

head(schooling_2023)

write_csv(
schooling_2023,
here("output", paste0("schooling-cleaned-", target_year, ".csv"))
)
```
## GDP per capita data
```{r}
lines <- readLines(path_gdp, n = 300, warn = FALSE)
header_line <- which(grepl("Country Name", lines, fixed = TRUE))[1]

gdp_raw <- read_csv(path_gdp, skip = header_line - 1, show_col_types = FALSE)

year_idx <- match(as.character(target_year), names(gdp_raw))
if (is.na(year_idx)) stop("Year column not found. Please check if this file contains the target year.")

gdp_country_2023 <- tibble(
country = gdp_raw[[1]],
iso_code = gdp_raw[[2]],
year = target_year,
gdp_per_capita = as.numeric(gdp_raw[[year_idx]])
) %>%
filter(nchar(iso_code) == 3)

head(gdp_country_2023)

write_csv(
gdp_country_2023,
here("output", paste0("gdp-per-capita-cleaned-", target_year, ".csv"))
)
```
# Data merging
```{r}
merged_yr <- fossil_share_elec_2023 %>%
left_join(
co2_per_capita_2023 %>% select(iso_code, year, co2_per_capita),
by = c("iso_code", "year")
) %>%
left_join(
schooling_2023 %>% select(iso_code, year, schooling),
by = c("iso_code", "year")
) %>%
left_join(
gdp_country_2023 %>% select(iso_code, year, gdp_per_capita),
by = c("iso_code", "year")
) %>%
drop_na(fossil_share_elec, co2_per_capita, schooling, gdp_per_capita)
head(merged_yr)

write_csv(
merged_yr,
here("output", paste0("merged_dataset_", target_year, "_complete_cases.csv"))
)
```
# Descriptive statistics
```{r}
df_desc <- merged_yr %>%
select(fossil_share_elec, co2_per_capita, schooling, gdp_per_capita) %>%
na.omit()

cat("Number of observations:", nrow(df_desc), "\n\n")

summary(df_desc)
```
## Boxplots
```{r}
op <- par(no.readonly = TRUE)
par(mfrow = c(2, 2), mar = c(4, 4, 5, 2), oma = c(0, 0, 2, 0))

boxplot(
df_desc$fossil_share_elec,
main = expression(bold(atop("Fossil Fuel Share of Electricity", "(%) Across Countries in 2023"))),
ylab = expression(bold("Fossil Fuel Share (%)")),
col = "lightblue"
)

boxplot(
df_desc$co2_per_capita,
main = expression(bold(atop(CO[2]~"Emissions per Capita", "(Tonnes per Capita) in 2023"))),
ylab = expression(bold(CO[2]~"Emissions (Tonnes)")),
col = "lightblue"
)

boxplot(
df_desc$schooling,
main = expression(bold(atop("Mean Years of Schooling", "(Years) Across Countries in 2023"))),
ylab = expression(bold("Schooling (Years)")),
col = "lightblue"
)

boxplot(
df_desc$gdp_per_capita,
main = expression(bold(atop("GDP per Capita", "(US$ per Capita) in 2023"))),
ylab = expression(bold("GDP per Capita (US$)")),
col = "lightblue"
)

mtext("Descriptive Statistics: Distribution of Variables", outer = TRUE, cex = 1.2, font = 2)
par(op)
```
## Correlation matrix
```{r}
cor_matrix <- round(
cor(df_desc, use = "pairwise.complete.obs"),
3
)

kable(cor_matrix, caption = "Correlation Matrix (2023)")
```
## Scatterplot matrix
```{r}
panel.lm <- function(x, y, ...) {
points(x, y, pch = 16, col = rgb(0.27, 0.51, 0.71, alpha = 0.6))
abline(lm(y ~ x), col = "red", lty = 2, lwd = 1.5)
}

pairs(
df_desc,
upper.panel = panel.lm,
lower.panel = panel.lm,
labels = c(
"Fossil Share\n(%)",
expression(CO[2]~"\n(Tonnes)"),
"Schooling\n(Years)",
"GDP per Cap\n(US$)"
),
main = "Scatterplot Matrix with Linear Trend Lines"
)
```

# Modelling
## Winsor function and model dataframe
```{r}
winsor <- function(x, probs = c(0.01, 0.99)) {
qs <- quantile(x, probs = probs, na.rm = TRUE, type = 7)
pmin(pmax(x, qs[1]), qs[2])
}

df_model <- merged_yr %>%
transmute(
fossil_share = as.numeric(fossil_share_elec),
co2_pc       = as.numeric(co2_per_capita),
schooling    = as.numeric(schooling),
gdp_pc       = as.numeric(gdp_per_capita)
) %>%
filter(complete.cases(fossil_share, co2_pc, schooling, gdp_pc)) %>%
mutate(
co2_pc_w = winsor(co2_pc, probs = c(0.01, 0.99)),
co2_shift = ifelse(min(co2_pc_w, na.rm = TRUE) <= 0,
abs(min(co2_pc_w, na.rm = TRUE)) + 1e-6, 0),
ln_co2_pc = log(co2_pc_w + co2_shift),
ln_gdp = log(gdp_pc),
fossil_share_c = fossil_share - mean(fossil_share, na.rm = TRUE),
schooling_c    = schooling    - mean(schooling, na.rm = TRUE)
)
```
## Three models
```{r}
m1 <- lm(ln_co2_pc ~ fossil_share_c + ln_gdp, data = df_model)
m2 <- lm(ln_co2_pc ~ fossil_share_c + ln_gdp + schooling_c, data = df_model)
m3 <- lm(ln_co2_pc ~ fossil_share_c + ln_gdp + schooling_c + fossil_share_c:schooling_c,data = df_model)
```
## Robust standard errors (HC1)
```{r}
se_m1_robust <- sqrt(diag(vcovHC(m1, type = "HC1")))[names(coef(m1))]
se_m2_robust <- sqrt(diag(vcovHC(m2, type = "HC1")))[names(coef(m2))]
se_m3_robust <- sqrt(diag(vcovHC(m3, type = "HC1")))[names(coef(m3))]
```
## Regression table
```{r stargazer-table, results='asis', echo=TRUE}
stargazer(
  m1, m2, m3,
  type = "html",
  title = "<span style='color: black; font-weight: bold;'> Table 2. Regression results of three models",
  dep.var.labels = "log(CO2 emissions per capita)",
  se = list(se_m1_robust, se_m2_robust, se_m3_robust),
  digits = 3,
  model.numbers = TRUE,
  order = c("fossil_share_c", "ln_gdp", "schooling_c", "fossil_share_c:schooling_c"),
  covariate.labels = c(
    "Fossil fuel share of electricity (centered)",
    "log(GDP per capita)",
    "Average years of schooling (centered)",
    "Fossil share × Schooling"
  ),
  notes.append = FALSE,
  notes = "Significance levels: * p < 0.05; ** p < 0.01; *** p < 0.001. <br> Robust standard errors (HC1) in parentheses.",
  header = FALSE
)
```
## Simple slope analysis
```{r simple-slopes, echo=TRUE}
s_slopes <- sim_slopes(
  model = m3,
  pred  = "fossil_share_c",
  modx  = "schooling_c",
  johnson_neyman = TRUE,
  cond.int = TRUE,
  robust = TRUE
)
print(s_slopes)

cat("### Table 3a: Simple Slopes (4-digit precision)\n")
s_slopes$slopes %>%
  as.data.frame() %>%
  kable(digits = 4, caption = "Slopes of fossil_share_c at different schooling levels")
```
## Interaction Plot
```{r interaction-plot, echo=TRUE, fig.height=7, fig.width=9}
interact_plot(
  model = m3,
  pred  = "fossil_share_c",
  modx  = "schooling_c",
  interval = TRUE,
  plot.points = TRUE,
  point.alpha = 0.3,
  x.label = "Fossil fuel share of electricity (centered)",
  y.label = "log(CO2 emissions per capita)",
  main.title = "Interaction: Schooling x Fossil Share (log CO2)"
)
```

# Model diagnostics
```{r}
par(mfrow = c(2, 2), mar = c(4.5, 4.5, 3, 2), oma = c(0, 0, 2, 0))
plot(
m3,
pch = 20,
col = rgb(120/255, 170/255, 215/255, alpha = 0.7),
lwd = 1.5,
sub.caption = ""
)
mtext("Regression Diagnostics for Moderation Model (log CO2)", outer = TRUE, cex = 1.2, font = 2)
par(mfrow = c(1, 1))
```
# VIF and model comparison
```{r}
vif_m3 <- vif(m3)
vif_m2 <- vif(m2)
vif_comparison <- data.frame(
Variable = names(vif_m3),
VIF_with_Interaction = as.numeric(vif_m3),
VIF_without_Interaction = as.numeric(vif_m2[names(vif_m3)])
)

kable(vif_comparison, digits = 3, caption = "VIF Comparison (with vs without interaction)")
```
# F-test (Model 2 vs Model 3)
```{r}
f_test_m2_m3 <- anova(m2, m3)
kable(f_test_m2_m3, caption = "F-test: Compare Model 2 vs Model 3")
```

